---
title: "Pre-processing kinobeads data"
author: "Junyan Lu"
date: "2021-10-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Load packages

Packages
```{r message=FALSE, warning=FALSE}
library(readxl)
library(tidyverse)
source("../code/utils.R")
```

# Read data sets
The drug-target data can be found in the supplementary file of the paper (Table_S1 & Table_S2):
https://science.sciencemag.org/content/358/6367/eaan4368/tab-figures-data


```{r warning=FALSE}
# kinase inhibitor drug-target screen from supplementary tables (Table_S1 & Table_S2)
tarList <- read_xlsx("../data/kinobeads/Klaeger_allTargets.xlsx", sheet = "Kinobeads")
tarList_syn <- read_xlsx("../data/kinobeads/targets_pubchem.xlsx", sheet = "Inhibitor annotation", col_names = TRUE) 
```

# Preprocess dataset

Attach synonyms to target table
```{r}
tarList$synonyms <- tarList_syn$Synonyms[match(tarList$Drug, tarList_syn$Drug)]
```

Process drug names of target table
```{r}
#use lowercase
tarList <- mutate(tarList, Drug = tolower(Drug)) %>%
  mutate(Drug = gsub("[- ]","", Drug))

#change pd325901 to pd0325901 (due to different naming)
tarList[tarList$Drug == "pd325901",]$Drug <- "pd0325901"
```

Rename BCR to BCR/ABL to avoid confusion with B-cell receptor (BCR)
```{r}
tarList <- mutate(tarList, `Gene Name` = ifelse(`Gene Name` %in% "BCR", "BCR/ABL", `Gene Name`))
```

# Plot distributions of drugs and targe kinases

### Kinases per drug
```{r}
plotTab <- tarList %>% filter(`Target Classification` =="High confidence") %>%
  group_by(Drug) %>% summarise(nTar = length(`Gene Name`))

p1 <- ggplot(plotTab, aes(x=nTar)) + geom_histogram(alpha=0.5, col="grey20") + 
  scale_x_continuous(n.breaks = 10) +
  xlab("number of kinases") + ylab("count") +
  xlim(0,100) + ylim(0,50) +
  ggtitle("Distribution of number \nof kinase targets per drug") +
  theme_custom
```

### Drug per kinases
```{r}
plotTab <- tarList %>% filter(`Target Classification` =="High confidence") %>%
  group_by(`Gene Name`) %>% summarise(nDrug = length(Drug))

p2 <- ggplot(plotTab, aes(x=nDrug)) + geom_histogram(alpha=0.5, col="grey20") + 
  scale_x_continuous(n.breaks = 10) +
  xlab("number of drugs") + ylab("count") +
  xlim(0,100) + ylim(0,70) +
  ggtitle("Distribution of number of \ndrugs per kinase target") +
  theme_custom
```


### Combine for Figure1
```{r Number_target_drug_distribution, fig.width=16, fig.height=6, dev=c("png","pdf") }
cowplot::plot_grid(p1,NULL, p2, rel_widths = c(1,0.1,1), ncol=3)
#ggsave("test.pdf", height = 5, width = 12)
```

# Save kinobeads data
```{r}
saveRDS(tarList, file = "../output/allTargets.rds")
```

